---
title: "Introducing simulations of heterogeneous treatments effects using the SPRINT population"
author: "CarDS Lab @Yale; Oikonomou EK, Khera R et al"
date: "8/6/2022"
output:
  html_document: default
---

## Introduction
This is R code to create a reproducible dataset from the BioLINCC SPRINT dataset for simulation analyses.
First, we will load the required libraries that we will use for preprocessing.
We also need to set a random seed number to ensure reproducibility (at the level of the imputation)

```{r}
# clean the environment
rm(list=ls()) 

# load the required libraries that we will use for preprocessing
library(plyr)
library(dplyr)
library(missRanger)
library(Hmisc)
library(janitor)
library(simsurv)
library(Publish)
library(survival)

# set a random seed
set.seed(123)
```

## Creating the dataset of baseline features
Next, for every file of interest, we will load the respective csv file and extract the information of interest.
The working directory/path should be adjusted to match the local path to the SPRINT data.

```{r}

# set the working directory to the file containing all the original .csv files
mypath = "/Users/evangelos.oikonomou/Box/Evan Rohan shared folder/Datasets/All datasets/"
setwd(file.path(mypath, "SPRINT_2021a/SPRINT/data/CSV"))

## Baseline blood pressure measurements
basebp <- read.csv("bp_manage_base.csv", header = TRUE, sep =",", na.strings=".") # baseline bp
basebp <- basebp %>% select("MASKID", "SEATSYS", "SEATDIAST", "SEATHEART", "STANDSYS", "STANDDIAST", "STANDHEART", "DIZZY")
basebp$SYS_CHANGE <- basebp$SEATSYS-basebp$STANDSYS
basebp$DIAST_CHANGE <- basebp$SEATDIAST-basebp$STANDDIAST
basebp$HR_CHANGE <- basebp$SEATHEART-basebp$STANDHEART
basebp[-c(1, 8)] <- lapply(basebp[-c(1,8)], as.numeric)
basebp[c(1, 8)] <- lapply(basebp[c(1,8)], as.factor)

## medical history
history <- read.csv("bl_history.csv", header = TRUE, sep =",", na.strings=".") # history
history <- history %>% select("MASKID",
                              "ATRIALFIB", "ANGINA", "HEARTATT", "CONHEARTFAIL", "IRRHEARTBEAT", 
                              "OSTEOARTHRITIS", "RHEARTHRITIS", "GOUT", "OTHARTHRITIS", 
                              "HIPPROB", 
                              "CANCER", "SKINCANCER", "PVD", "SEIZURE", "STROKE", "TIA", 
                              "THYROIDDIS", "ANEMIA", "DIABETES", "HYPERTENS", "LOWBKPAIN", 
                              "ALCOHOL",
                              "FAMHST", "FAMHST55", 
                              "VIGACTIV")

history$FAMHST <- ifelse(history$FAMHST==2, NA, history$FAMHST) # define missing values
history$FAMHST55 <- ifelse(history$FAMHST55==2, NA, history$FAMHST55) # define missing values

history[2:26] <- lapply(history[2:26], factor) # factorize all selected variables

## Cognitive testing
mind <- read.csv("bl_mindscreening.csv", header = TRUE, sep =",", na.strings=".") # mind
mind <- mind %>% select("MASKID", "MOCA_SCORE", "DSC_TOTAL")
mind[2:3] <- lapply(mind[2:3], as.numeric)

## Electrocardiograms
ecg <- read.csv("ecg.csv", header = TRUE, sep =",", na.strings=".") # ecg
ecg <- ecg %>% filter(INCI_MC=="")
ecg <- ecg %>% group_by(MASKID) %>% filter(row_number()==1)
ecg <- ecg %>% select("MASKID", "QRSDURATION",
                      "AFIBFLUTTER",
                      "CV", "CVP", "SL", 
                      "LVHANY3")   
ecg[-c(1, 3, 7)] <- lapply(ecg[-c(1, 3, 7)], as.numeric)
ecg[c(1, 3, 7)] <- lapply(ecg[c(1, 3, 7)], as.factor)

## Inclusion/excusion criteria
inclexcl <- read.csv("incl_excl.csv", header = TRUE, sep =",", na.strings=".") # inclusion/exclusion
inclexcl <- inclexcl %>% select("MASKID",         
                                "MYOCARDINFARC", "ACUTECORSYND", "CORONARYREVAS", "CAROTID", "PADREVAS", "STENOSIS50", "AAA5REPAIR",
                                "CALCSCORE400", "LOWABI90", 
                                "UNABLETOSTAND",
                                "CVDPOINTS")
inclexcl[-c(12)] <- lapply(inclexcl[-c(12)], factor)
inclexcl[c(12)] <- lapply(inclexcl[c(12)], as.numeric)
inclexcl$UNABLETOSTAND <- as.factor(ifelse(is.na(inclexcl$UNABLETOSTAND), 0, 1))

## Laboratory assessment
labs <- read.csv("labs.csv", header = TRUE, sep =",", na.strings=".") # labs
labs <- labs %>% filter(VISITCODE=="RZ1")
labs <- labs %>% select("MASKID", "RESULT_BUN", "RESULT_CHR", "RESULT_CL", "RESULT_CO2", "RESULT_CRDUR", "RESULT_GLUR", "RESULT_HDL", "RESULT_K", "RESULT_LDLR",
                        "RESULT_NA", "RESULT_TRR", "RESULT_UMALCR", "RESULT_UMALI", "RESULT_CREATR", "RESULT_GFR")
labs[-c(1)] <- lapply(labs[-c(1)], as.numeric)

## General self-reported health assessment
myhealth <- read.csv("my_health.csv", header = TRUE, sep =",", na.strings=".") # myhealth
myhealth <- myhealth %>% filter(VISITCODE=="RZ2")
myhealth <- myhealth %>% select("MASKID", "GEN_HEALTH", "FAINT")
myhealth[-c(2)] <- lapply(myhealth[-c(2)], as.factor)
myhealth[c(2)] <- lapply(myhealth[c(2)], as.numeric)

# load selected files from POP
setwd(file.path(mypath, "SPRINT_2021a/SPRINT-POP/data"))
baseline <- read.csv("baseline.csv", header = TRUE, sep =",", na.strings=".") # baseline
baseline <- baseline %>% select("MASKID", "INTENSIVE", "AGE", "FEMALE", "RACE4", "BMI", "N_AGENTS", "NOAGENTS", "SMOKE_3CAT", "ASPIRIN", "STATIN", "SUB_CVD", "SUB_CKD")
baseline$SMOKE_3CAT <- ifelse(baseline$SMOKE_3CAT==4, NA, baseline$SMOKE_3CAT)
baseline[c(3, 6, 7)] <- lapply(baseline[c(3,6,7)], as.numeric)
baseline[-c(3, 6, 7)] <- lapply(baseline[-c(3,6,7)], as.factor)
```

## Baseline data imputation
We will combine the data and then imputed missing values using chained random forests with predictive mean matching deployed within the missRanger package. Prior to imputation we can exlclude variables with greater than e.g. 10% missing data.

```{r}
## Base(line) phenotypes
base <- join_all(list(baseline, inclexcl, history, myhealth, basebp, labs, ecg, mind), by='MASKID', type='left')
arm <- base$INTENSIVE
MASKID <- base$MASKID
base <- base[,-c(1,2)]

# Impute missing data

# keep only variables with e.g. <10% missing data
base <- base[colSums(is.na(base))/nrow(base)<0.1]

# missing data imputation
imputed <- missRanger(base, maxiter=1L)

# we preprocess the names using the clean_names function of the janitor package
imputed <- clean_names(imputed)
```

## Introducing heterogeneous treatment effects (HTE)
Now we will introduce a ground truth for HTE. For this, we use the method proposed in the study: https://trialsjournal.biomedcentral.com/articles/10.1186/s13063-018-2774-5
This based on the code availabe through: https://github.com/joerigdon/HTE.

First, a function is defined to create a multinomial distribution for a given variable with defined probabilities.

```{r}
rmulti = function(prob, n) {
  as.numeric(rMultinom(probs=matrix(prob, 1, length(prob)), m=n))
}
```

## Defining groups with HTE.
Next, we will define responders for the HTE. e.g. we assume that women on aspirin are responders to the treatment.

```{r}
## Define responders and non-responders for the HTE
imputed$delta = NA #simulate delta
responder = c(imputed$female==1 & imputed$aspirin==1)
nonresponder = c(imputed$female!=1 | imputed$aspirin!=1)
```

## Defining the size of the HTE.
We then need to introduce the expected effect size among "responders" and "non-responders".
For instance, if there is HTE in our population; the effect can be introduced with the following code:

```{r}
## (i) Simulation for HTE
imputed$delta[responder] = rmulti(c(0.09, 0.88, 0.03), length(imputed$delta[responder]))-2
mean(imputed$delta[responder]) 

imputed$delta[nonresponder] = rmulti(c(0.06, 0.88, 0.06), length(imputed$delta[nonresponder]))-2
mean(imputed$delta[nonresponder])

```

Alternatively, in the absence of HTE, the code can look like this:

```{r}
## (ii) Simulation without HTE
imputed$delta = rmulti(c(0.065, 0.88, 0.055), length(imputed$delta))-2
mean(imputed$delta) 
```

The average ATE (avrage treatment effect) in the population can be estimated with the following code:

```{r}
mean(imputed$delta)
```

Next, we need to create the simulated treatment and outcome variables;

Z: study arm (0: control, 1: intervention)
Y1: potential adverse medical event when randomized to treatment
Y0: potential adverse medical event when randomized to placebo
Y: the outcome for each individual was calculated as: Z ? Y1 + (1 ? Z) ? Y0 

```{r}
## Now we will create the treatment and outcome variables
imputed$Y1 = NA
imputed$Y0 = NA
imputed$Y1[imputed$delta==-1] = 0
imputed$Y0[imputed$delta==-1] = 1
imputed$Y1[imputed$delta==1] = 1
imputed$Y0[imputed$delta==1] = 0
imputed$Y1[imputed$delta==0] = 0
imputed$Y0[imputed$delta==0] = 0
sum((imputed$Y1-imputed$Y0)!=imputed$delta) #they all equal delta
mean(imputed$Y1-imputed$Y0) # as specified above for ATE
mean(imputed$delta)
mean(imputed$Y1)
mean(imputed$Y0)

##Add in rowname
imputed$id = paste(rownames(imputed))

##Change types where necessary
imputed$delta = factor(imputed$delta)
imputed$Y1 = factor(imputed$Y1)
imputed$Y0 = factor(imputed$Y0)
```

We also need to create random treatment allocation and random events times - in this case we assume a Gompertz distribution:

```{r}
covs <- data.frame(id = 1:nrow(imputed), trt = stats::rbinom(nrow(imputed), 1L, 0.5))
s2 <- simsurv(dist = "gompertz", lambdas = 0.1, gammas = 0.05, x = covs)
imputed$Z <- covs$trt

imputed$Y = NA
imputed$Y[imputed$Z==1] = imputed$Y1[imputed$Z==1]
imputed$Y[imputed$Z==0] = imputed$Y0[imputed$Z==0]
imputed$Y = imputed$Y-1
imputed$TIME <- s2$eventtime
```

We can check the data to confirm the presence of HTE:

```{r}
# identifying factors for analysis
imputed$Z <- as.factor(imputed$Z)
imputed$group <- as.factor(ifelse((responder), "responder", "non-responder"))

# fitting cox regression model
fit <- coxph(Surv(TIME, Y) ~ Z + group, data=imputed)
fit
sub_cox <- subgroupAnalysis(fit, data=imputed, treatment="Z", subgroups=c("group"))
sub_cox
plot(sub_cox)
```